Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures

### 2020-10-09 manuscript/manuscript.tex
R1 for jrn

\documentclass[12pt,twocolumn,twoside,]{pinp}

@@

\usepackage[switch]{lineno} \linenumbers \usepackage{setspace} \onehalfspacing

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Mokken scale analysis of the perioperative thirst discomfort
scale}

@@

% initially we use doi so keep for backwards compatibility
% new name is doi_footer
\doifooter{Pre-print submitted to \emph{Applied Nursing Research}}

@@

Pre-procedure fasting is used to reduce the risk of vomiting and
aspiration pneumonia during sedation and general anaesthesia, or in case
emergency intubation is required due to unexpected cardiac arrest
\citep{hamid2014pre, osborne2002preoperative}. However, prolonged fluid
restriction causes thirst symptoms to develop (e.g., dry mouth, swollen
tongue), which can lead to great discomfort
\citep{madsen1998perioperative}. Current guidelines related to
pre-procedure fasting for elective procedures recommend a minimum
fasting period of 2 hours nil-per-os (NPO) for clear fluids
\citep{dobsonGuidelinesPracticeAnesthesia2018}. Despite these
recommendations, current practice is for patients undergoing surgical
and other medical procedures that require sedation or anesthesia to
receive standardized `nil-by-mouth' fasting instructions at a
pre-specified time interval before procedures. For example, `no eating
or drinking after midnight' is most common. This `standardized'
instruction will not be changed regardless of whether or not there are
alterations in scheduling throughout the day that result in significant
delays in procedure start time. As a result, fasting durations far
exceed the recommended requirement for most patients undergoing medical
and surgical procedures
\citep{de2014actual, sorita2015frequency, spitz2017impact}. For example,
in a recent study of 3641 fasting orders at a large academic institution
in the USA, it was found that the median fasting duration was 12.8
hours, averaging 2 missed meals \citep{sorita2015frequency}.

@@

Despite the relevance and value of assessing thirst-discomfort of
patients, the subjective experience of thirst presents challenges in
developing a valid and reliable tool to succinctly and accurately
measure its symptoms and level of discomfort prior to procedures. A
thirst-discomfort scale for perioperative use was developed at a
surgical centre affiliated with an accredited public hospital in Brazil
\citep{Martins_2017}. The seven-item perioperative thirst-discomfort
scale (PTDS) was developed in three stages, including face and content
validation and reliability, and based on the Consensus-based Standards
for the selection of health status Measurement Instruments (COSMIN)
checklist \citep{mokkink2010cosmin}. Inter-rater reliability was tested
through inter-observer equivalence where a pair of nurses independently,
but simultaneously administered the scale among 70 patients. Six items
on the scale had a weighted kappa coefficient of 1, while the item ``I
feel like drinking water'' was 0.97 \citep{Martins_2017}. Cronbach's
alpha was used to test the internal consistency of the scale with a
value of 0.91 \citep{Martins_2017}.

Although the initial validation of the PTDS was conducted in a different
context (i.e.~\emph{post}-surgical patients), it demonstrates strong
potential for accurately assessing thirst-discomfort more generally for
peri-operative use. The aim of this study was to examine the validity
and reliability of the PTDS for measuring thirst discomfort for patients
who are fasting \emph{before} medical procedures.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Under 16 years of age
\item
  Scheduled for an emergency procedure
\item
  Unable to understand or speak English and a translator was not
  immediately available
\item
  Nurse in pre-procedure bay considered that there was insufficient time
  prior to anticipated commencement of the procedure for participation
  in the study.
\end{enumerate}

@@

The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a
7-item self-reported, composite score evaluating the severity of dry
mouth, lips, and throat, thick tongue and saliva, a bad taste in the
mouth, and a desire to drink water. The total score can range from 0 to
14, with 14 representing the ``most extreme'' intensity of discomfort
related to perioperative thirst. This instrument was developed and
validated for perioperative use in a surgical centre at a public
hospital in Brazil. Face validation was conducted by group consisting of
nurses with PhDs, new graduate nurses, and patients. Content validation
was conducted by a team consisting of specialists with perioperative
nursing experience related to thirst or instrument validation. The
reliability of the scale was assessed through inter-observer equivalence
by a research nurse and a staff nurse from the Post Anesthesia Care Unit
among 70 patients pre and postoperatively in an anesthetic recovery
room. The Cronbach alpha score was 0.91. The time taken to complete the
PTDS-7 was not evaluated.

@@

The R statistical program was used to conduct all analyses
\citep{rcore2020}. All data and R code, along with instructions for how
to completely reproduce results of the analyses, is available
\href{https://github.com/awconway/ptds}{here}.

@@

The distribution of PTDS-5 scores in the sample was right-skewed, with a
median score of 3 (IQR 1, 6). (Figure \ref{fig:distributionpdf}). As
shown in Figure \ref{fig:globalthirst}, there was a positive correlation
between PTDS-5 scores and the global thirst discomfort rating
(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63). The correlation between
PTDS-5 scores and thirst intensity was also high (\emph{tau} = 0.49;
95\% CI = 0.38 to 0.59). Likewise, the correlation between pain and
PTDS-5 scores was not significant (Spearman's rho = 0.07; 95\% CI =
-0.07 to 0.21).

@@

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/distributionpdf-1}

@@

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/globalthirst-1}

@@

{\centering \includegraphics[width=252px]{manuscript_files/figure-latex/fasting-1}

@@

Results indicate that the PTDS-5 is a reliable and precise way to order
patients according to their thirst discomfort in the
\emph{pre-procedure} period. In addition, an acceptable accuracy for
invariant item ordering was identified, meaning patients are likely to
rank the PTDS-5 symptoms of thirst discomfort in the same way. As such,
the item difficulty hierarchy, presented in Table
\ref{tab:hierarchytabpdf}, can be used as a guide to easily determine
which patients are experiencing worse thirst discomfort. The
hierarchical pattern of responses to items in the Mokken scale is
interpretable in terms of respondents more easily endorsing items
related to a general desire to drink water through to first endorsing
more specific symptoms related to dryness of the mouth and lips prior to
endorsing discomfort associated with the sensations of `thick' saliva
and `thick' tongue. In more practical terms, a patient who reports
feeling bothered by a sensation that saliva is thick, could be
interpreted as experiencing worse thirst discomfort than a patient who
did not endorse any of the items lower down on the item difficulty
hierarchy (i.e.~`my lips are dry' and `my mouth is dry').

@@

We did not observe an association between PTDS-5 and fasting duration
for either food or fluids. This would suggest that it is important to
assess thirst discomfort periodically, regardless of the duration of
fasting. If thirst discomfort is present, and fasting is still
indicated, interventions to alleviate symptoms should be implemented.
There is a lack of evidence regarding the efficacy of interventions to
alleviate perioperative thirst discomfort. However, recent randomized
controlled trials have identified clinically relevant redictions in
thirst discomfort with menthol chewing gum and 30mL menthol popsicles
\citep{garcia2019, nascimento2020}.

@@

Selection bias should be considered because a convenience sample was
used for this study. Also, it is uncertain if similar results would be
obtained for patients who are fasted prior to undergoing procedures in
other settings. Only one patient was excluded from the analysis due to
missing data, so it is unlikely attrition bias would have exerted a
major influence on the results. Further research is required to evaluate
the minimal detectable difference for the PTDS-5 as well as the minimal
clinically important difference prior to use of this scale as an outcome
measure in research.

@@

The items in the PTDS-5 form a reliable and precise way to order
patients according to their thirst discomfort. Nurses may use this scale
to assess the severity of thirst discomfort in patients who are fasted
prior to procedures and initiate interventions according to individual
needs.

>>>>

\documentclass[12pt,twoside,]{pinp}

@@

\usepackage{setspace} \onehalfspacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Validating the perioperative thirst discomfort scale for
measuring thirst discomfort prior to procedures}

@@

% initially we use doi so keep for backwards compatibility
% new name is doi_footer

@@

Pre-procedure fasting is used to reduce the risk of vomiting and
aspiration pneumonia during sedation and general anaesthesia, or in case
emergency intubation is required due to unexpected cardiac arrest
\citep{hamid2014pre, osborne2002preoperative}. However, prolonged fluid
restriction causes thirst symptoms to develop (e.g., dry mouth, swollen
tongue), which can lead to great discomfort
\citep{madsen1998perioperative}. Current guidelines related to
pre-procedure fasting for elective procedures recommend a minimum
fasting period of 2 hours nil-per-os (NPO) for clear fluids
\citep{dobsonGuidelinesPracticeAnesthesia2018}. Despite these
recommendations, current practice is for patients undergoing surgical
and other medical procedures that require sedation or anesthesia to
receive standardized `nil-by-mouth' fasting instructions at a
pre-specified time interval before procedures. For example, `no eating
or drinking after midnight' is most common. It is not common for fasting
instructions to be updated even when there are significant delays in
procedure start time. As a result, fasting durations far exceed the
recommended requirement for most patients undergoing medical and
surgical procedures
\citep{de2014actual, sorita2015frequency, spitz2017impact}. For example,
in a recent study of 3641 fasting orders at a large academic institution
in the USA, it was found that the median fasting duration was 12.8
hours, averaging 2 missed meals \citep{sorita2015frequency}.

@@

Despite the relevance and value of assessing thirst-discomfort of
patients, the subjective experience of thirst presents challenges in
developing a valid and reliable tool to succinctly and accurately
measure its symptoms and level of discomfort prior to procedures. A
thirst-discomfort scale for perioperative use was developed at a
surgical centre affiliated with an accredited public hospital in Brazil
\citep{Martins_2017}. The seven-item perioperative thirst-discomfort
scale (PTDS) was developed in three stages, including face and content
validation and reliability, and based on the Consensus-based Standards
for the selection of health status Measurement Instruments (COSMIN)
checklist \citep{mokkink2010cosmin}. Inter-rater reliability was tested
through inter-observer equivalence where a pair of nurses independently,
but simultaneously administered the scale among 70 patients. Six items
on the scale had a weighted kappa coefficient of 1, while the item ``I
feel like drinking water'' was 0.97 \citep{Martins_2017}. Cronbach's
alpha was used to test the internal consistency of the scale with a
value of 0.91 \citep{Martins_2017}. It should be noted that the initial
validation of the PTDS was conducted in one specific point in the
perioperative journey (i.e.~\emph{post}-surgical patients). The
psychometric properties of the scale may be influenced by events that
occur during the surgical procedure, such as fluid loss and
administrations of medications that induce thirst, such as
glycopyrrolate. A recent large study showed that use of glycopyrrolate
was an independent risk factor for moderate to severe post-operative
thirst.\citep{lee2020prevalence} As such, validation of the scale for
measuring thirst discomfort before procedures is required.

The PTDS demonstrates strong potential for accurately assessing
thirst-discomfort more generally for peri-operative use. The aim of this
study was to examine the validity and reliability of the PTDS for
measuring thirst discomfort for patients who are fasting \emph{before}
medical procedures.

@@

Patients were not eligible for the study if they were under 16 years of
age, scheduled for an emergency procedure, unable to understand or speak
English (unless a translator was available) or if the nurse in
pre-procedure bay considered that there was insufficient time prior to
anticipated commencement of the procedure for participation in the
study.

@@

The seven-item Perioperative Thirst Discomfort Scale (PTDS-7) is a
7-item self-reported, composite score evaluating the severity of dry
mouth, lips, and throat, thick tongue and saliva, a bad taste in the
mouth, and a desire to drink water.\citep{Martins_2017} The total score
can range from 0 to 14, with 14 representing the ``most extreme''
intensity of discomfort related to perioperative
thirst.\citep{Martins_2017}

@@

The R statistical program was used to conduct all analyses
\citep{rcore2020}. All data and R code, along with instructions for how
to completely reproduce results of the analyses, is available
\href{https://github.com/awconway/ptds}{here}. A formal sample size
calculation was not conducted.

@@

The distribution of PTDS-5 scores in the sample was right-skewed, with a
median score of 3 (IQR 1, 6). (Figure \ref{fig:distributionpdf}). As
shown in Figure \ref{fig:globalthirst}, there was a positive correlation
between PTDS-5 scores and the global thirst discomfort rating
(\emph{tau} = 0.54; 95\% CI = 0.43 to 0.63; 0). The correlation between
PTDS-5 scores and thirst intensity was also high (\emph{tau} = 0.49;
95\% CI = 0.38 to 0.59). Likewise, the correlation between pain and
PTDS-5 scores was not significant (Spearman's rho = 0.07; 95\% CI =
-0.07 to 0.21).

@@

{\centering \includegraphics[width=252px]{distributionpdf-1}

@@

{\centering \includegraphics[width=252px]{globalthirst-1}

@@

{\centering \includegraphics[width=252px]{fasting-1}

@@

Results indicate that the PTDS-5 is a reliable and precise way to order
patients according to their thirst discomfort in the
\emph{pre-procedure} period. In addition, an acceptable accuracy for
invariant item ordering was identified, meaning patients are likely to
rank the PTDS-5 symptoms of thirst discomfort in the same way. As such,
the item difficulty hierarchy, presented in Table
\ref{tab:hierarchytabpdf}, can be used as a guide to easily determine
which patients are experiencing worse thirst discomfort. The
hierarchical pattern of responses to items in the Mokken scale is
interpretable in terms of respondents more easily endorsing items
related to a general desire to drink water through to first endorsing
more specific symptoms related to dryness of the mouth and lips prior to
endorsing discomfort associated with the sensations of `thick' saliva
and `thick' tongue. In more practical terms, a patient who reports feeling bothered by a sensation of thick saliva could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e.~`my lips are dry' and `my mouth is dry').

@@

We did not observe an association between PTDS-5 and fasting duration
for either food or fluids. This would suggest that it is important to
assess thirst discomfort periodically, regardless of the duration of
fasting. If thirst discomfort is present, and fasting is still
indicated, interventions to alleviate symptoms should be implemented.
There is a lack of evidence regarding the efficacy of interventions to
alleviate perioperative thirst discomfort. However, recent randomized
controlled trials have identified clinically relevant reductions in
thirst discomfort with menthol chewing gum and 30mL menthol popsicles
\citep{garcia2019, nascimento2020}. One additional potential course of
action could be to rearrange the order of procedures on a schedule based
on results of periodic thirst discomfort assessments using the PTDS. For
example, circumstance may arise where a patient who reports severe
thirst discomfort is scheduled to have their procedure later in the day
than a patient who is not currently bothered by thirst symptoms.
Reordering the procedure schedule would reduce the duration of fasting
for the patient who is bothered by symptoms of thirst. Further research
would be required to confirm the effectiveness of this strategy in
practice on overall patient satisfaction.

@@

Selection bias should be considered because a convenience sample was
used for this study. Also, it is uncertain if similar results would be
obtained for patients who are fasted prior to undergoing procedures in
other settings. Only one patient was excluded from the analysis due to
missing data, so it is unlikely attrition bias would have exerted a
major influence on the results. Further research is required to evaluate
the minimal detectable difference for the PTDS-5 as well as the minimal
clinically important difference prior to use of this scale as an outcome
measure in research. The potential that the sample size was insufficient
for Mokken scale analysis should be considered. A recent study
empirically evaluated the sample size required for a scale and found
that a minimum of 250 participants would be required to establish
scalability of the whole scale.\citep{watson2018} It was unclear,
however, if this minimum threshold would apply to other scales.

@@

The items in the PTDS-5 may provide a reliable and precise way to order
patients according to their thirst discomfort. Nurses may use this scale
to assess the severity of thirst discomfort in patients who are fasted
prior to procedures and initiate interventions according to individual
needs. The PTDS-5 with the rating scale is presented in Table
\ref{tab:tabptds}.

\caption{\label{tab:tabptds}5-item perioperative thirst discomfort scale (PTDS-5)}
\centering
\begin{tabular}[t]{lrrr}
\toprule
How bothered are you by the following? & Not at all & Slightly & Extremely\\
\midrule
I want to drink water & 0 & 1 & 2\\
My mouth is dry & 0 & 1 & 2\\
My lips is dry & 0 & 1 & 2\\
My saliva is thick & 0 & 1 & 2\\
My tongue is thick & 0 & 1 & 2\\
\bottomrule
\end{tabular}
\end{table}
